---
title: "R Notebook"
output: html_notebook
---


```{r}
weight=read.csv("C:\\Users\\agelf\\Documents\\School\\U of C\\Masters of Data Science and Analytics\\Data 603\\Group Project\\weight_change_dataset.csv")
colnames(weight)
```
```{r}
library(GGally)
```

```{r}
weight_plot=data.frame(weight$Age,weight$BMR..Calories.,weight$Daily.Calories.Consumed,weight$Daily.Caloric.Surplus.Deficit,weight$Weight.Change..lbs.,weight$Duration..weeks.)
ggpairs(weight_plot,lower = list(continuous = "smooth_loess", combo =
  "facethist", discrete = "facetbar", na = "na"))
```

Full model:

#```{r}
#weight_model_full=lm(Weight.Change..lbs.~Age+factor(Gender)+Current.Weight..lbs.+BMR..Calories.+Daily.Calories.Consumed+Daily.Caloric.#Surplus.Deficit+Duration..weeks.+factor(Physical.Activity.Level)+factor(Sleep.Quality)+factor(Stress.Level)+Final.Weight..lbs., #data=weight)
#summary(weight_model_full)
#```

REMOVE CURRENT WEIGHT AND FINAL WEIGHT

```{r}
weight_model_take2=lm(Weight.Change..lbs.~Age+factor(Gender)+BMR..Calories.+Daily.Calories.Consumed+Daily.Caloric.Surplus.Deficit+Duration..weeks.+factor(Physical.Activity.Level)+factor(Sleep.Quality)+factor(Stress.Level), data=weight)
summary(weight_model_take2)
```
```{r}
alias(weight_model_take2)
```
```{r}
vif(weight_model_take2)
imcdiag(weight_model_take2, method="VIF")
```
REMOVE BMR AND DAILY CALORIES CONSUMED, AS DAILY CALORIC SURPLUS/DEFICIT IS CALCULATED FROM THE 2

```{r}
weight_model_take3=lm(Weight.Change..lbs.~Daily.Caloric.Surplus.Deficit+Age+factor(Gender)+Duration..weeks.+factor(Physical.Activity.Level)+factor(Sleep.Quality)+factor(Stress.Level), data=weight)
summary(weight_model_take3)
```
KEEP DURATION, SLEEP QUALITY, STRESS LEVEL

```{r}
vif(weight_model_take3)
imcdiag(weight_model_take3, method="VIF")
```

```{r}
library(olsrr)
```


```{r}
weight_model_take4=lm(Weight.Change..lbs.~factor(Sleep.Quality)+Duration..weeks.+factor(Stress.Level), data=weight)
summary(weight_model_take4)
```

CHECK FOR INT TERMS
```{r}
weight_model_take4_int=lm(Weight.Change..lbs.~(factor(Sleep.Quality)+Duration..weeks.+factor(Stress.Level))^2, data=weight)
summary(weight_model_take4_int)
```
KEEP SLEEP, DURATION, STRESS, SLEEPxDURATION, DURATIONxSTRESS

```{r}
weight_model_take4_int_2=lm(Weight.Change..lbs.~factor(Sleep.Quality)+Duration..weeks.+factor(Stress.Level)+factor(Sleep.Quality)*Duration..weeks.+Duration..weeks.*factor(Stress.Level), data=weight)
summary(weight_model_take4_int_2)
```
```{r}
plot(weight_model_take4_int_2)
```

```{r}
library(Ecdat)
library(lmtest)
bptest(weight_model_take4_int_2)
shapiro.test(residuals(weight_model_take4_int_2))
```
FAILS ASSUMPTIONS


```{r}
library(MASS) #for the boxcox()function
bc=boxcox(weight_model_take4_int_2,lambda=seq(-5,5)) #CAN USE (-5,5) for lambda
#extract best lambda
bestlambda=bc$x[which(bc$y==max(bc$y))]
bestlambda
```
CAN'T RUN BOXCOX DUE TO NEGATIVE RESPONSE,



```{r}
weight_model_take4_int_2_2=lm((Weight.Change..lbs.+40)~factor(Sleep.Quality)+Duration..weeks.+factor(Stress.Level)+factor(Sleep.Quality)*Duration..weeks.+Duration..weeks.*factor(Stress.Level), data=weight)
summary(weight_model_take4_int_2_2)
```


```{r}
library(MASS) #for the boxcox()function
bc=boxcox(weight_model_take4_int_2_2,lambda=seq(-5,5)) #CAN USE (-5,5) for lambda
#extract best lambda
bestlambda=bc$x[which(bc$y==max(bc$y))]
bestlambda
```

```{r}
bptest(weight_model_take4_int_2_2)
shapiro.test(residuals(weight_model_take4_int_2_2))
```

```{r}
adjusted_model=lm(((((Weight.Change..lbs.+40)^2.373737)-1)/2.373737)~factor(Sleep.Quality)+Duration..weeks.+factor(Stress.Level)+factor(Sleep.Quality)*Duration..weeks.+Duration..weeks.*factor(Stress.Level), data=weight)
summary(adjusted_model)
```

```{r}
bptest(adjusted_model)
shapiro.test(residuals(adjusted_model))
```


IGNORE THIS STUFF FOR NOW

```{r}
weight2=weight
```


```{r}
weight2$Gender=as.factor(weight2$Gender)
weight2$Physical.Activity.Level=as.factor(weight2$Physical.Activity.Level)
weight2$Sleep.Quality=as.factor(weight2$Sleep.Quality)
weight2$Stress.Level=as.factor(weight2$Stress.Level)
```

```{r}
weight_model_take4=lm(Weight.Change..lbs.~Age+Gender+BMR..Calories.+Daily.Calories.Consumed+Daily.Caloric.Surplus.Deficit+Duration..weeks.+Physical.Activity.Level+Sleep.Quality+Stress.Level, data=weight2)
summary(weight_model_take4)
```

```{r}
weight_step2=ols_step_both_p(weight_model_take4,p_enter = 0.05, p_remove = 0.06, details=TRUE)
summary(weight_step2$model)
```

```{r}
weight_take5=lm(Weight.Change..lbs.~Sleep.Quality+Stress.Level+Duration..weeks.+Daily.Caloric.Surplus.Deficit,data=weight2)
summary(weight_take5)
```

Use the figures
```{r}
library(GGally)
```

```{r}
weightfigs = data.frame(weight2$Weight.Change..lbs.,weight2$Sleep.Quality,weight2$Stress.Level,weight2$Duration..weeks.,weight2$Daily.Caloric.Surplus.Deficit)
#LOESS or LOWESS: LOcally WEighted Scatter-plot Smoother
ggpairs(weightfigs,lower = list(continuous = "smooth_loess", combo = "facethist", discrete = "facetbar", na = "na"))
#pairs(~Y+X1+X2+factor(X3)+X4+X5,data=salary,panel = panel.smooth) 
```


```{r}
weight_take5_int=lm(Weight.Change..lbs.~(Sleep.Quality+Stress.Level+Duration..weeks.+Daily.Caloric.Surplus.Deficit)^2,data=weight2)
summary(weight_take5_int)
```
```{r}
weight_take5_int_reduced=lm(Weight.Change..lbs.~Sleep.Quality + Stress.Level + Duration..weeks. + Daily.Caloric.Surplus.Deficit + Sleep.Quality*Stress.Level + Stress.Level*Duration..weeks. + Stress.Level*Daily.Caloric.Surplus.Deficit,data=weight2)
summary(weight_take5_int_reduced)
```
```{r}
anova(weight_take5_int_reduced,weight_take5_int)
```

```{r}
anova(weight_take5_int_reduced,weight_take5)
```

```{r}
alias(weight_take5_int_reduced)
```


```{r}
imcdiag(weight_take5_int_reduced, method="VIF")
```

