---
title: "DATA 603 L01 05 Final Report Appendix"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
---
title: "DATA 603 L01 05 Final Report Appendix"
output: html_notebook
---
Group L01-05 Members:

1. Steen Rasmussen - 30097313
2. Aaron Gelfand - 10032214
3. Jackson Meier - 30095291
4. Harpreet Saini - 30271048
5. Venkateshwaran Balu Soundararajan - 30239509

# 1 Appendix

## 1.1 Final Report Code

Below we have included all the code, while hiding the output, as all of the output was previously presented in the final report.

```{r, results='hide'}
library(ggplot2)
library(GGally)
library(car)
library(lmtest)
library(MASS)
library(mctest)
library(leaps)
library(olsrr)
library(Ecdat)
weight=read.csv("https://raw.githubusercontent.com/aarongelf/Data-603-Group-Project/refs/heads/main/weight_change_dataset.csv")
print(head(weight))
```

### 1.1.1 Quantitative Model

```{r, results='hide'}
colnames(weight)
```

```{r, results='hide'}
quant_weight_model_full = lm(Weight.Change..lbs. ~ Age + `Current.Weight..lbs.` + `BMR..Calories.` + `Daily.Calories.Consumed` + `Daily.Caloric.Surplus.Deficit` + Duration..weeks. + `Final.Weight..lbs.`, data = weight)

summary(quant_weight_model_full)
```

```{r, results='hide'}
quant_weight_model_take2 = lm(Weight.Change..lbs. ~ Age + `BMR..Calories.` + 
                         `Daily.Calories.Consumed` + `Daily.Caloric.Surplus.Deficit` + Duration..weeks., data = weight)
summary(quant_weight_model_take2)
```

```{r, results='hide'}
vif(quant_weight_model_take2)
imcdiag(quant_weight_model_take2, method="VIF")
```

```{r, results='hide'}
quant_weight_model_take3 = lm(Weight.Change..lbs. ~ Age + `Daily.Caloric.Surplus.Deficit` + Duration..weeks., data = weight)
summary(quant_weight_model_take3)
```

**NOTE** The code line is hidden here, as when left to run it produces errors, preventing the document from knitting. This was to simply demonstrate that when attempting to use stepwise selection, none of our predictors meet the criteria to be included in the model.

```{r, results='hide'}
#stepmod=ols_step_both_p(quant_weight_model_take3, p_enter=0.05,p_remove=0.3,details=TRUE)
```

### 1.1.2 Qualitative Model

```{r, results='hide'}
print(head(weight))
```

```{r, results='hide'}
Weight_full_model= lm(Weight.Change..lbs.~ factor(Gender)+factor(Physical.Activity.Level)+factor(Sleep.Quality)+factor(Stress.Level), data=weight)
print(summary(Weight_full_model))
```

```{r, results='hide'}
weight_refined_model=lm(Weight.Change..lbs.~factor(Sleep.Quality)+factor(Stress.Level), data=weight)
print(summary(weight_refined_model))
```

```{r, results='hide'}
vif(weight_refined_model)
imcdiag(weight_refined_model,method='VIF')
```

```{r, results='hide'}
weight_interaction_model= lm(Weight.Change..lbs.~(factor(Sleep.Quality)+factor(Stress.Level))^2, data=weight)
print(summary(weight_interaction_model))
```

```{r, results='hide'}
weight_higher_order_model = lm(Weight.Change..lbs. ~ (factor(Sleep.Quality) + factor(Stress.Level))^3, data=weight)
print(summary(weight_higher_order_model))
```

```{r, results='hide'}
# Residuals vs Fitted Values
ggplot(weight_refined_model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_smooth(method = "loess", color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals")
```

```{r, results='hide'}
plot(residuals(weight_refined_model), type = "o", col = "blue", 
     main = "Residuals vs Observation Number",
     xlab = "Observation Number", ylab = "Residuals")
abline(h = 0, lty = 2, col = "red")
```

```{r, results='hide'}
ggplot(weight_refined_model, aes(x = .fitted, y = .resid)) +
  geom_point(color = "purple") +
  geom_smooth(method = "loess", color = "green4") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals")

# Breusch-Pagan Test
library(lmtest)
bptest(weight_interaction_model)
```

```{r, results='hide'}
# QQ Plot
qqnorm(residuals(weight_refined_model))
qqline(residuals(weight_refined_model), col = "red")

# Shapiro-Wilk Test
shapiro.test(residuals(weight_refined_model))
```

```{r, results='hide'}
# Cook's Distance
plot(weight_refined_model, which = 4, col = "red", pch = 18, 
     main = "Cook's Distance Plot")

# Observations with high Cook's distance
weight[cooks.distance(weight_refined_model) > 0.5, ]

# Leverage values
lev = hatvalues(weight_refined_model)
p = length(coef(weight_refined_model))
n = nrow(weight)
outliers_2pn = lev[lev > (2 * p / n)]
outliers_3pn = lev[lev > (3 * p / n)]

# Leverage plot
plot(1:n, lev, type = "h", main = "Leverage Plot", 
     xlab = "Observation Index", ylab = "Leverage")
abline(h = c(2 * p / n, 3 * p / n), col = c("blue", "red"), lty = 2)
legend("topright", legend = c("2p/n", "3p/n"), col = c("blue", "red"), lty = 2)
```

```{r, results='hide'}
print(summary(weight_refined_model))
```

### 1.1.3 Model Including All Variables

```{r, results='hide'}
weight=read.csv("https://raw.githubusercontent.com/aarongelf/Data-603-Group-Project/refs/heads/main/weight_change_dataset.csv")
head(weight)
```

```{r, results='hide'}
weight_model_full=lm(Weight.Change..lbs. ~ Age + factor(Gender) + BMR..Calories. + Daily.Calories.Consumed + Daily.Caloric.Surplus.Deficit + Duration..weeks. + factor(Physical.Activity.Level) + factor(Sleep.Quality) + factor(Stress.Level), data=weight)
print(summary(weight_model_full))
```

```{r, results='hide'}
vif(weight_model_full)
imcdiag(weight_model_full, method="VIF")
```

```{r, results='hide'}
weight_model_adj=lm(Weight.Change..lbs.~Daily.Caloric.Surplus.Deficit+Age+factor(Gender)+Duration..weeks.+factor(Physical.Activity.Level)+factor(Sleep.Quality)+factor(Stress.Level), data=weight)
summary(weight_model_adj)
```

```{r, results='hide'}
weight_model_adj2=lm(Weight.Change..lbs.~factor(Sleep.Quality)+Duration..weeks.+factor(Stress.Level), data=weight)
summary(weight_model_adj2)
```

```{r, results='hide'}
vif(weight_model_adj2)
imcdiag(weight_model_adj2, method="VIF")
```

```{r, results='hide'}
weight_model_int=lm(Weight.Change..lbs.~(factor(Sleep.Quality)+Duration..weeks.+factor(Stress.Level))^2, data=weight)
summary(weight_model_int)
```

```{r, results='hide'}
weight_model_int_final=lm(Weight.Change..lbs.~factor(Sleep.Quality)+Duration..weeks.+factor(Stress.Level)+factor(Sleep.Quality)*Duration..weeks.+Duration..weeks.*factor(Stress.Level), data=weight)
summary(weight_model_int_final)
```

```{r, results='hide'}
ggplot(weight_model_int_final, aes(x=.fitted, y=.resid)) +
geom_point() +geom_smooth()+
geom_hline(yintercept = 0)
```

```{r, results='hide'}
weight_model_poly=lm(Weight.Change..lbs.~factor(Sleep.Quality)+Duration..weeks.+I(Duration..weeks.^2)+factor(Stress.Level)+factor(Sleep.Quality)*Duration..weeks.+Duration..weeks.*factor(Stress.Level), data=weight)
summary(weight_model_poly)
```

```{r, results='hide'}
ggplot(weight_model_poly, aes(x=.fitted, y=.resid)) +
geom_point() +geom_smooth()+
geom_hline(yintercept = 0)
```

```{r, results='hide'}
weight_model_log <- lm(Weight.Change..lbs. ~ factor(Sleep.Quality) + log(Duration..weeks.) + factor(Stress.Level) + factor(Sleep.Quality) * log(Duration..weeks.) + log(Duration..weeks.) * factor(Stress.Level), data = weight)
summary(weight_model_log )
```

```{r, results='hide'}
ggplot(weight_model_log, aes(x=.fitted, y=.resid)) +
geom_point() +geom_smooth()+
geom_hline(yintercept = 0)
```

```{r, results='hide'}
plot(weight_model_int_final,which=1)
plot(weight_model_int_final,which=3)
bptest(weight_model_int_final)
```

```{r, results='hide'}
plot(weight_model_log,which=1)
plot(weight_model_log,which=3)
bptest(weight_model_log)
```

```{r, results='hide'}
plot(weight_model_log,which=2)
shapiro.test(residuals(weight_model_log))
```

```{r, results='hide'}
plot(weight_model_log,which=5)
```

```{r, results='hide'}
weight_clean=weight[!weight$Participant.ID==49,]
weight_model_log_cleaned=lm(Weight.Change..lbs. ~ factor(Sleep.Quality) + log(Duration..weeks.) + factor(Stress.Level) + factor(Sleep.Quality) * log(Duration..weeks.) + log(Duration..weeks.) * factor(Stress.Level), data = weight_clean)
summary(weight_model_log_cleaned)
```

```{r, results='hide'}
new_weight_data=data.frame(Sleep.Quality='Poor',Stress.Level=5,Duration..weeks.=6)
predict(weight_model_log_cleaned,new_weight_data,type="response")
```

## 1.2 Code for Stepwise Selection for Model Including All Variables

To determine the best model for predicting weight change, given our dataset, we begin by uploading the data set.
```{r}
weight=read.csv("https://raw.githubusercontent.com/aarongelf/Data-603-Group-Project/refs/heads/main/weight_change_dataset.csv")
colnames(weight)
```
```{r}
weight$Stress.Level[which(weight$Stress.Level==1)]="D1"
weight$Stress.Level[which(weight$Stress.Level==2)]="D2"
weight$Stress.Level[which(weight$Stress.Level==3)]="D3"
weight$Stress.Level[which(weight$Stress.Level==4)]="D4"
weight$Stress.Level[which(weight$Stress.Level==5)]="D5"
weight$Stress.Level[which(weight$Stress.Level==6)]="D6"
weight$Stress.Level[which(weight$Stress.Level==7)]="D7"
weight$Stress.Level[which(weight$Stress.Level==8)]="D8"
weight$Stress.Level[which(weight$Stress.Level==9)]="D9"
head(weight)
```

We then make sure all libraries needed are uploaded
```{r}
library(GGally)
library(mctest)
library(car)
library(ggplot2)
library(Ecdat)
library(lmtest)
library(olsrr)
library(leaps)
```
```{r}
weight_model_full=lm(Weight.Change..lbs. ~ Age + Gender + Daily.Caloric.Surplus.Deficit + Duration..weeks. + Physical.Activity.Level + Sleep.Quality + Stress.Level, data=weight)
summary(weight_model_full)
```

We then attempt a stepwise selection method to determine the best predictors
```{r}
stepmod=ols_step_both_p(weight_model_full, p_enter=0.05,p_remove=0.1,details=TRUE)
```

```{r}
ExecSubsets=ols_step_best_subset(weight_model_full, details=TRUE) # for the output interpretation 
ExecSubsets$metrics
```



```{r}
rsquare=c((ExecSubsets$metrics)$rsquare) 
rse=c((ExecSubsets$metrics)$rse)
AdjustedR=c((ExecSubsets$metrics)$adjr)
cp=c((ExecSubsets$metrics)$cp) 
AIC=c((ExecSubsets$metrics)$aic) 
cbind(rsquare,AdjustedR,rse,cp,AIC)
```

From the stepwise selection method, the adjust $R^2$, the cp, and the AIC, we can see that the model with 4 variables should be selected, which includes Daily.Caloric.Surplus.Deficit, Duration..weeks., Sleep.Quality, and Stress.Level. However, for reasons discussed in our final report, we choose to select our predictors using the t-test results from our full model summary.