---
title: "Qualitative variables"
author: "Harpreet"
date: "2024-11-13"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Required Libraries

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(car)
library(lmtest)
library(MASS)
library(mctest)
library(leaps)
library(olsrr)
```

# Dataset Loading.

```{r}

weight=read.csv("D:\\RData603\\Project\\weight_change_dataset.csv")
print(colnames(weight))
print(head(weight))
```

Meet our, 100 participants who has decided to embark on a transformative journey to improve their health and well-being. Over the next 12 weeks, they are determined to contribute to body test which is going to be really helpful in the health department to understand how various lifestyle factors influence the body weight.

Starting down with the following variables -

1.  Gender- M or F
2.  Duration..weeks.
3.  Physical.Activity.Level
4.  Sleep.Quality
5.  Stress.Level

Both genders commits to a 12-week program to track their progress with how does their physical activity level is throughout that period of time. Not only the physical monitoring is done also the mental aspects were kept under radar such as there sleep quality and stress level was also monitered.

$H_0: \beta_1 = \beta_2 =...= \beta_i=0\mbox{    (i=1,2,...,p)}$
$H_a: \mbox{ At least one } \beta_i \neq 0\mbox{    (i=1,2,...,p)}$

```{r}
weight_plot <- data.frame(Duration = factor(weight$Duration..weeks.),
  Activity = factor(weight$Physical.Activity.Level),
  Sleep = factor(weight$Sleep.Quality),
  Stress = factor(weight$Stress.Level),
  Gender = factor(weight$Gender)
)

# Create the pair plot using categorical options
ggpairs(weight_plot, 
        lower = list(combo = "facethist", discrete = "facetbar", na = "na"))

```


#Full Model

```{r}
Weight_full_model= lm(Weight.Change..lbs.~ factor(Gender)+ factor(Duration..weeks.)+factor(Physical.Activity.Level)+factor(Sleep.Quality)+factor(Stress.Level), data=weight)
print(summary(Weight_full_model))
```

$$\begin{multline*}
\text{Weight.Change (lbs)} = 5.367 - 6.686 \cdot \text{factor(Duration..weeks.)7} 
- 5.807 \cdot \text{factor(Duration..weeks.)8} - 5.790 \cdot \text{factor(Duration..weeks.)9} \\
- 5.993 \cdot \text{factor(Duration..weeks.)10} - 11.635 \cdot \text{factor(Duration..weeks.)11} 
- 4.304 \cdot \text{factor(Duration..weeks.)12} - 7.245 \cdot \text{factor(Sleep.Quality)Poor} \\
- 10.663 \cdot \text{factor(Stress.Level)8} - 9.861 \cdot \text{factor(Stress.Level)9}
\end{multline*}$$

The most impactfull factors for weight change are specific durations (weeks 7–12), poor sleep quality, and very high stress levels (8 and 9), also the adjusted R Squared 0.7478 indicates the model explains approximately 75% of the variability in weight change, with these predictors.


# So removing the Physical.Activity.Level and Gender because they were non significant.
```{r}
weight_refined_model=lm(Weight.Change..lbs.~factor(Duration..weeks.)+factor(Sleep.Quality)+factor(Stress.Level), data=weight)
print(summary(weight_refined_model))
```
$$\begin{align*}
\text{Weight.Change (lbs)} &= 5.1303 - 5.0981 \cdot \text{factor(Duration..weeks.)7} \\
&\quad - 5.3606 \cdot \text{factor(Duration..weeks.)8} - 5.2530 \cdot \text{factor(Duration..weeks.)9} \\
&\quad - 5.3338 \cdot \text{factor(Duration..weeks.)10} - 10.3586 \cdot \text{factor(Duration..weeks.)11} \\
&\quad - 4.0025 \cdot \text{factor(Duration..weeks.)12} - 7.0672 \cdot \text{factor(Sleep.Quality)Poor} \\
&\quad - 11.6240 \cdot \text{factor(Stress.Level)8} - 10.1985 \cdot \text{factor(Stress.Level)9}.
\end{align*}$$



# Calculate Variance Inflation Factor (VIF)

```{r}

vif(weight_refined_model) # under library car
imcdiag(weight_refined_model,method='VIF')

```
As we are checking is If VIF > 5 consider removing or combining variables.
But in our case the VIF is less than 5 so all three predictors are acceptable(we are keep all 3) as NO significant multicollinearity.

Refined model is good.

# Interaction Model- with centered variables 
#the next step is to assess whether adding interaction terms improves the model. Interaction terms are useful when we suspect that the effect of one predictor on the response variable depends on the level of another predictor.

```{r}

weight_interaction_model= lm(Weight.Change..lbs.~(factor(Duration..weeks.)+factor(Sleep.Quality)+factor(Stress.Level))^2, data=weight)
print(summary(weight_interaction_model))

```

Main Effects: Weight change is significantly influenced by duration, with notable losses at weeks 8, 9, and 11, and by stress levels, particularly stress level 3.

Interaction Effects: Significant interactions between duration and sleep quality (especially in week 11) show large weight changes, while week 2 and stress level 3 also indicate a significant weight change.


```{r}
weight_interaction_model2= lm(Weight.Change..lbs.~factor(Duration..weeks.)+factor(Sleep.Quality)+factor(Stress.Level)+
                                Duration..weeks.* Sleep.Quality + Duration..weeks.*Stress.Level , data=weight)
print(summary(weight_interaction_model2))
```


The first interactive model has a significantly higher R-squared value, indicating it explains a higher proportion of the variance in the data. However, we might consider the second model because it includes the interaction terms, which are important for capturing complex relationships between the variables (e.g., between duration, stress level, and sleep quality).


# Higher order model on Interction model 1 as the R squre is higher in that than interaction model 2
Notes-Use of I(factor()^2) is for continuous variables where you want to square the variable, but it's not applicable to categorical variables.
As here are all categorical variables we will just do it like - 

```{r}
weight_higher_order_model = lm(Weight.Change..lbs. ~ (factor(Duration..weeks.) + factor(Sleep.Quality) + factor(Stress.Level))^3, data=weight)
print(summary(weight_higher_order_model))

```
Oh nice its same , as by doing the higher order as well it gives the same values which means we can stick to our interaction model.


# Residuals - check code
```{r}
plot(weight_interaction_model)
```





# Linearity Assumption - check code
```{r}

# Residuals vs Fitted Values
ggplot(weight_interaction_model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_smooth(method = "loess", color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals")

```


# Independence Assumption - check code
```{r}
plot(residuals(weight_interaction_model), type = "o", col = "blue", 
     main = "Residuals vs Observation Number",
     xlab = "Observation Number", ylab = "Residuals")
abline(h = 0, lty = 2, col = "red")
```

# Equal Variance Assumption - check code

```{r}
ggplot(weight_interaction_model, aes(x = .fitted, y = .resid)) +
  geom_point(color = "purple") +
  geom_smooth(method = "loess", color = "green4") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals")

# Breusch-Pagan Test
bptest(weight_interaction_model)
```
Since the p-value is much smaller than 0.05, we reject the null hypothesis (which assumes homoscedasticity which means that there is statistically significant evidence of heteroscedasticity in the model.the residuals indicates that the current set of predictors may not fully explain the variability in the response variable, and additional variables could potentially reduce or eliminate the heteroscedasticity.

# normality
```{r}
# QQ Plot
qqnorm(residuals(weight_interaction_model))
qqline(residuals(weight_interaction_model), col = "red")

# Shapiro-Wilk Test
shapiro.test(residuals(weight_interaction_model))
```

the Shapiro-Wilk test shows that the residuals (the differences between the predicted and actual values) from your model are not normally distributed. This means that the model might not meet one of the key assumptions for accurate statistical testing, which could affect the reliability of the results, such as p-values and confidence intervals.


# Check for Outliers - check code
```{r}
# Cook's Distance
plot(weight_interaction_model, which = 4, col = "red", pch = 18, 
     main = "Cook's Distance Plot")

# Observations with high Cook's distance
weight[cooks.distance(weight_interaction_model) > 0.5, ]

# Leverage values
lev = hatvalues(weight_interaction_model)
p = length(coef(weight_interaction_model))
n = nrow(weight)
outliers_2pn = lev[lev > (2 * p / n)]
outliers_3pn = lev[lev > (3 * p / n)]

# Leverage plot
plot(1:n, lev, type = "h", main = "Leverage Plot", 
     xlab = "Observation Index", ylab = "Leverage")
abline(h = c(2 * p / n, 3 * p / n), col = c("blue", "red"), lty = 2)
legend("topright", legend = c("2p/n", "3p/n"), col = c("blue", "red"), lty = 2)
```
The leverage plot suggests that there are no major concerns with influential observations. All observations contribute fairly equally to the regression model. This is a positive finding for the reliability and stability of the model.


Perform the ANOVA test between the selected interaction model (as the full model) and the reduced model (as the simpler model).
Note: To avoid the overfitting  issue we will be choosing the First Interactive model 1 as it also shows the higher Adj R Square and Lower RSE than the Interaction model 2. 

Model 2, with its interaction effects, may give you more insights into how predictors work together but at the cost of interpretability, but our  goal is to quickly understand the influence of each predictor without delving into complex interactions. 

# To have the final model

# anova test
```{r}
print(anova(weight_refined_model,weight_interaction_model))
```
The p-value for the F-statistic (0.0819) suggests that there is weak evidence that the interaction terms in Model 2 significantly improve the model compared to refined model. This suggests that while the interaction terms do explain some additional variation, the difference between the two models is not statistically significant at the 0.05 level, though it might be at the 0.1 level.

So We decided based on this ANOVA result that Refined model  (with only main effects- Duration..weeks, Sleep.Quality and Stress.Level seems to be a better model, as the additional complexity of interaction terms does not significantly enhance the explanatory power of the model at the typical significance level (0.05).

Final Model :

```{r}
print(summary(weight_interaction_model))
```


$$\begin{align*}
\text{Weight.Change (lbs)} &= 5.1303 - 5.0981 \cdot \text{factor(Duration..weeks.)7} \\
&\quad - 5.3606 \cdot \text{factor(Duration..weeks.)8} - 5.2530 \cdot \text{factor(Duration..weeks.)9} \\
&\quad - 5.3338 \cdot \text{factor(Duration..weeks.)10} - 10.3586 \cdot \text{factor(Duration..weeks.)11} \\
&\quad - 4.0025 \cdot \text{factor(Duration..weeks.)12} - 7.0672 \cdot \text{factor(Sleep.Quality)Poor} \\
&\quad - 11.6240 \cdot \text{factor(Stress.Level)8} - 10.1985 \cdot \text{factor(Stress.Level)9}.
\end{align*}$$

In conclusion, our final model explains the relationship between weight change under three key factors: Duration of weeks, Sleep Quality, and Stress Level. The analysis shows that several levels of Duration and Stress Level have significant effects on weight change, while Sleep Quality does not show strong evidence of a consistent impact. Specifically, we observe that Stress Level has significant negative effects at higher levels (e.g., Stress Level 8 and 9), while Duration has a negative impact on weight change as the duration increase.